---
output: github_document
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# add2compareGroups

Extends the `compareGroups` package to seamlessly add adjusted **Odds Ratios (OR)** and **Hazard Ratios (HR)** into descriptive tables — perfect for real-world data analysis where raw and adjusted measures must coexist in reports. Also generates **Forest Plots** for visual comparison.

🌱 **Work in progress!** We’re actively improving this package. Feedback and suggestions are welcome! 

## Installation

Install  the devolopment version from GitHub:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("mrocalou/add2compareGroups")
```


## Full tutorial (vignette)

All examples, use cases and detailed explanations are available in the package vignette. Once installed, run:

```{r, eval = FALSE}
browseVignettes("add2compareGroups")
```

Or view it directly in RStudio’s Help pane after loading the package.
